---
title: "State Service Los Frequency of Use Table"
author: "Connor Flynn"
date: "12/29/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(janitor)
library(haven)
library(feather)
library(arrow)
#library(chorddiag)
library(circlize)
library(kableExtra)
library(urbnmapr)
library(forcats)
library(sf)
library(tidycensus)
options(scipen = 99)
```

Load in Data

```{r}
teds_d_select_years <- arrow::read_feather(here("data/teds_d_15_19.feather"))
```



<!-- 14 million treatment episodes from 2012-2020 -->

<!-- ```{r} -->
<!-- teds_d <- arrow::read_feather(here("data/tedsD_2012_2020.feather")) -->
<!-- ``` -->


<!-- Clean Column Names -->

<!-- ```{r} -->
<!-- teds_d <- teds_d %>%  -->
<!--   clean_names() -->
<!-- ``` -->


<!-- Filter for years with admission/discharge data (2015-2020, 9 million records) -->


<!-- ```{r} -->
<!-- teds_d_select_years <- teds_d %>%  -->
<!--   filter(disyr %in% c("2015", "2016", "2017", "2018", "2019", "2020")) -->
<!-- ``` -->

<!-- Count records per year -->

<!-- ```{r} -->
<!-- table(teds_d_select_years$disyr) -->
<!-- ``` -->

*Filter for Completed Treatments ONLY* (4 million treatment episodes)

```{r}
teds_d_select_years_completed <- teds_d_select_years %>%
filter(reason == "1")
```


<!-- Replace -9 with NA -->

<!-- ```{r} -->
<!-- teds_d_select_years_completed[teds_d_select_years_completed == "-9"] <- NA -->
<!-- ``` -->


Relevant variables to decrease size of large dataframe

```{r}
teds_d_select_years_completed_relevant <- teds_d_select_years_completed %>% 
  select(stfips, freq1_d, services_d, los)
```


Bin LOS
- easier for data visualization
- binned to limit number of classes and not bias random forest

```{r}
teds_d_select_years_completed_relevant$los_binned <- cut(
  teds_d_select_years_completed_relevant$los,
  breaks = c(0, 1, 5, 10, 20, 30, 31, 32, 33, 34, 35, 36, 37),
  labels = c('1 day', '2-5 days', '6-10 days', '11-20 days', '21-30 days', '31-45 days', '46-60 days', '61-90 days', '91-120 days', '121-180 days', '181-365', '>365 days'),
  include.lowest = TRUE
)

# Check the frequency table of the new variable
teds_d_select_years_completed_relevant %>% 
  select(los, los_binned)
```
```{r}
#teds_d_select_years_completed_relevant[is.na(teds_d_select_years_completed_relevant)] <- "unknown"
```

Replace frequency of use / services codes with meaning

```{r}
teds_d_select_years_completed_relevant$freq1_d[teds_d_select_years_completed_relevant$freq1_d == "1"] <- "no use"


teds_d_select_years_completed_relevant$freq1_d[teds_d_select_years_completed_relevant$freq1_d == "2"] <- "some use"


teds_d_select_years_completed_relevant$freq1_d[teds_d_select_years_completed_relevant$freq1_d == "3"] <- "daily use"


teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "1"] <- "Detox, 24-hour, hospital inpatient"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "2"] <- "Detox, 24-hour, free-standing residential"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "3"] <- "Rehab/residential, hospital (non-detox)"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "4"] <- "Rehab/residential, short term (30 days or fewer)"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "5"] <- "Rehab/residential, long term (more than 30 days)"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "6"] <- "Ambulatory, intensive outpatient"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "7"] <- "Ambulatory, non-intensive outpatient"

teds_d_select_years_completed_relevant$services_d[teds_d_select_years_completed_relevant$services_d == "8"] <- "Ambulatory, detoxification"


```



Frequency of Use 

```{r}
frequency_of_use_by_state <- teds_d_select_years_completed_relevant %>% 
  group_by(stfips, freq1_d) %>% 
  count(name = "sum_freq_use_in_state")
```


Mutate Percent column 
- each state will have four rows for freq1_d that should sum to 100%
```{r}
frequency_of_use_by_state <- frequency_of_use_by_state %>% 
  group_by(stfips) %>% 
  mutate(percent_freq_use = sum_freq_use_in_state / sum(sum_freq_use_in_state) * 100)
```

Drop sum since it was only used to calculate the percentage
- change state and frequency of use to factors

```{r}
frequency_of_use_by_state <- frequency_of_use_by_state %>% 
  select(-sum_freq_use_in_state) 


frequency_of_use_by_state$stfips <- as.factor(frequency_of_use_by_state$stfips)
frequency_of_use_by_state$freq1_d <- as.factor(frequency_of_use_by_state$freq1_d)
```


```{r}
# frequency_of_use_by_state_wider <- frequency_of_use_by_state %>% 
#   pivot_wider(names_from = freq1_d, values_from = percent_freq_use)
```



Services by State

```{r}
services_by_state <- teds_d_select_years_completed_relevant %>% 
  group_by(stfips, services_d) %>% 
  count(name = "sum_service_in_state")
```

Mutate Percent column 
- each state will have some number of rows for each service (whichever are offered in state) that should sum to 100%

```{r}
services_by_state <- services_by_state %>% 
  group_by(stfips) %>% 
  mutate(percent_in_service = sum_service_in_state / sum(sum_service_in_state) * 100)
```

Drop sum since it was only used to calculate the percentage
- change state and services at discharge to factors

```{r}
services_by_state <- services_by_state %>% 
  select(-sum_service_in_state) 


services_by_state$stfips <- as.factor(services_by_state$stfips)
services_by_state$services_d <- as.factor(services_by_state$services_d)
```

```{r}
# services_by_state_wider <- services_by_state %>% 
#   pivot_wider(names_from = services_d, values_from = percent_in_service)
```


LOS by State 

```{r}
los_by_state <- teds_d_select_years_completed_relevant %>% 
  group_by(stfips, los_binned) %>% 
  count(name = "sum_los_binned_in_state")
```

Mutate Percent column 
- each state will have 12 rows for each los bin  that should sum to 100%

```{r}
los_by_state <- los_by_state %>% 
  group_by(stfips) %>% 
  mutate(percent_in_los_bin = sum_los_binned_in_state / sum(sum_los_binned_in_state) * 100)
```

Drop sum since it was only used to calculate the percentage
- change state and los bin to factors

```{r}
los_by_state <- los_by_state %>% 
  select(-sum_los_binned_in_state)

los_by_state$stfips <- as.factor(los_by_state$stfips)
los_by_state$los_binned <- as.factor(los_by_state$los_binned)
```

```{r}
# los_by_state_wider <- los_by_state %>% 
#   pivot_wider(names_from = los_binned, values_from = percent_in_los_bin)
```


```{r}
# head(los_by_state_wider)
# head(services_by_state_wider)
# head(frequency_of_use_by_state_wider)
```

```{r}
head(los_by_state)
head(services_by_state)
head(frequency_of_use_by_state)
```

Plot to show concept for final tile map

```{r}
ggplot(frequency_of_use_by_state, aes(x = freq1_d, y = stfips, fill = percent_freq_use)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal()
```


Join Data


```{r}

# Join the dataframes
df_merged <- los_by_state %>%
  full_join(services_by_state, by = "stfips", relationship = "many-to-many") %>%
  full_join(frequency_of_use_by_state, by = "stfips", relationship = "many-to-many")


```

```{r}
df_merged_longer <- df_merged %>% 
  pivot_longer(cols = c("los_binned", "freq1_d", "services_d"),
               names_to = "importance_feature",
               values_to = "desription")

df_merged_longer2 <- df_merged_longer %>%
  mutate(percentage = case_when(
    importance_feature == "los_binned" ~ percent_in_los_bin,
    importance_feature == "freq1_d" ~ percent_freq_use,
    importance_feature == "services_d" ~ percent_in_service,
    TRUE ~ NA_real_
  )) 


df_merged_longer2 <- df_merged_longer2 %>% 
  select(stfips, importance_feature, desription, percentage) %>% 
  distinct()



# df_merged_longer <- df_merged_longer %>% 
#   pivot_longer(cols = c("percent_in_los_bin", "percent_freq_use", "percent_in_service"),
#                #names_to = "importance_feature",
#                values_to = "percentage")
```


```{r}
head(df_merged_longer2)
```
```{r}
df_merged_longer2$percentage <- round(df_merged_longer2$percentage, digit = 0)
```

Add state names based on fips codes

```{r}
state_info <- data.frame(
  stfips = c(1, 2, 4, 5, 6, 8, 9, 10, 11, 12, 13, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 42, 44, 45, 46, 47, 48, 49, 50, 51, 55, 56, 72),
  State = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Wisconsin", "Wyoming", "Puerto Rico")
)

df_merged_longer2 <- merge(df_merged_longer2, state_info, by = "stfips", all.x = TRUE)
```

```{r}
df_merged_longer2 <- df_merged_longer2 %>%
  mutate(importance_feature = case_when(
    importance_feature == "services_d" ~ "service at discharge",
    importance_feature == "freq1_d" ~ "frequency of use at discharge",
    importance_feature == "los_binned" ~ "length of stay",
    TRUE ~ importance_feature
  ))
```

```{r}
df_merged_longer2 <- df_merged_longer2 %>%
  mutate(desription = case_when(
    importance_feature == "frequency of use at discharge" ~ factor(
      desription,
      levels = c("no use", "some use", "daily use", "NA")
    ),
    TRUE ~ desription # Keep other values unchanged
  ))

```


Final tile map 

Check plot to make sure it matches after joining
1) los_by_state
2) services_by state
3) frequency_of_use_by_state

```{r}
# Load the lemon package
library(lemon)

# Updated tile map with lemon::facet_rep_grid
tile_map <- ggplot(df_merged_longer2, aes(x = desription, y = State, fill = percentage)) +
  geom_tile() +
  lemon::facet_rep_grid(~importance_feature, scales = "free_x", repeat.tick.labels = "left") +
  theme_minimal() +
  scale_fill_viridis_c() +
  geom_text(aes(label = percentage), color = "white", size = 2) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 7, color = "black"),
    axis.text.y = element_text(size = 7, color = "black"),
    strip.placement = "outside",  # Place facet labels outside
    panel.spacing = unit(1, "lines")  # Add space between facets for better readability
  ) +
  labs(x = "", y = "State", fill = "% of Treatment Episodes")

tile_map
```

```{r}
ggsave(
  filename = here("plots", "tile_map_slide_ready.png"),
  plot = tile_map,
  device = "png",
  width = 16,  # Adjust width
  height = 9,  # Adjust height
  dpi = 300
)
```



Make interactive

```{r}
#ggplotly(tile_map)
```

State Analysis

```{r}
teds_d <- arrow::read_feather(here("data/teds_d_15_19.feather"))
```

```{r}
teds_d_completed <- teds_d %>% 
  filter(reason == 1)
```

```{r}
teds_d_state_counts <- teds_d_completed %>% 
  group_by(stfips) %>% 
  count(name = "count") %>% 
  ungroup() %>% 
  mutate(percentage = count / sum(count) * 100)
```

```{r}
state_pop <- get_estimates(
  geography = "state",
  product = "population",
  vintage = 2022,
  year = 2022
) %>%
  filter(variable == "POPESTIMATE") %>%
  select(NAME, value) %>%
  rename(State = NAME, population = value)

# 3. Add FIPS codes to join with your TEDS-D data
state_pop <- state_pop %>%
  mutate( # Match state name to FIPS
    percent_of_us = population / sum(population) * 100
  )

# 4. Merge with your TEDS-D counts
teds_d_state_counts <- teds_d_completed %>% 
  group_by(stfips) %>% 
  count(name = "treatmet_episodes") %>% 
  ungroup() %>% 
  mutate(percent_of_teds_d = treatmet_episodes / sum(treatmet_episodes) * 100) # ensure FIPS is zero-padded for join

teds_d_state_counts <- merge(teds_d_state_counts, state_info, by = "stfips")

teds_d_state_counts <- teds_d_state_counts %>%
  left_join(state_pop, by = "State")

teds_d_state_counts <- teds_d_state_counts %>%
  mutate(
    percent_of_teds_d = round(percent_of_teds_d, 2),
    percent_of_us = round(percent_of_us, 2)
  )
```


```{r}
library(ggplot2)
library(dplyr)
library(forcats)

# Prepare the data
pyramid_data <- teds_d_state_counts %>%
  mutate(
    percent_of_teds_d = -percent_of_teds_d,  # negative for left side
    State = fct_reorder(State, percent_of_us)  # order states
  )

pyramid_plot <- ggplot(pyramid_data) +
  # Left bar: % TEDS-D
  geom_col(aes(x = percent_of_teds_d, y = State), fill = "#1f77b4") +
  # Right bar: % US population
  geom_col(aes(x = percent_of_us, y = State), fill = "#ff7f0e") +
  
  # Percentage labels inside bars
  geom_text(aes(x = percent_of_teds_d, y = State, 
                label = paste0(abs(percent_of_teds_d), "%")),
            hjust = 1.8, color = "black", size = 3) +
  
  geom_text(aes(x = percent_of_us, y = State, 
                label = paste0(percent_of_us, "%")),
            hjust = -0.6, color = "black", size = 3) +

  # State labels at center
  geom_text(aes(x = 0, y = State, label = State),
            hjust = 0.5, size = 2.2, fontface = "bold") +
  
  # Customize axes
  scale_x_continuous(
    breaks = seq(-15, 15, by = 5),
    labels = function(x) paste0(abs(x), "%"),
    limits = c(-20, 20)
  ) +
  labs(
    x = "Percent of Total",
    y = NULL,
    title = "State-Level Comparison of TEDS-D Treatment Episodes vs Population",
    subtitle = "Left: % of TEDS-D | Center: State | Right: % of U.S. Population",
    caption = "Source: TEDS-D and U.S. Census Bureau (Vintage 2022)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold")
  )

pyramid_plot

```

```{r}
ggsave(
  filename = here("plots", "pyramid.png"),
  plot = pyramid_plot,
  device = "png",
  width = 16,  # Adjust width
  height = 9,  # Adjust height
  dpi = 300
)
```



Health Insurance by State



```{r}
stfips_hlthins <- teds_d_completed %>%
  mutate(
    hlthins = factor(hlthins),  # Convert to categorical
    stfips = factor(stfips)     # Convert to categorical
  ) %>%
  count(stfips, hlthins) %>%
  group_by(stfips) %>%  # Group by state
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()
```


```{r}
stfips_hlthins <- merge(stfips_hlthins, state_info, by = "stfips")
```

```{r}
stfips_hlthins <- stfips_hlthins %>%
  mutate(hlthins = case_when(
    hlthins == "1" ~ "Private insurance, Blue Cross/Blue Shield, HMO",
    hlthins == "2" ~ "Medicaid",
    hlthins == "3" ~ "Medicare, other (e.g. TRICARE, CHAMPUS)",
    hlthins == "4" ~ "None",
    TRUE ~ hlthins
  ))
```

```{r}
medicaid_order <- stfips_hlthins %>%
  filter(hlthins == "Medicaid") %>%
  arrange(percentage) %>%
  pull(State)

stfips_hlthins <- stfips_hlthins %>%
  mutate(State = factor(State, levels = medicaid_order))
```



```{r}
state_hlthins_plot <- ggplot(stfips_hlthins, aes(x = State, y = percentage, fill = hlthins)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(
    aes(label = ifelse(percentage > 10, sprintf("%.1f%%", percentage), "")),
    position = position_stack(vjust = 0.5),
    size = 3,
    color = "black"
  ) +
  scale_fill_viridis_d(direction = -1) +
  theme_bw() +
  coord_flip() +
  ylim(0, 101)

state_hlthins_plot

```



```{r}
ggsave(
  filename = here("plots", "state_hlthins_plot.png"),
  plot = state_hlthins_plot,
  device = "png",
  width = 16,  # Adjust width
  height = 9,  # Adjust height
  dpi = 300
)
```


```{r}
teds_d_completed <- teds_d_completed %>%
  mutate(stfips = as.integer(stfips)) %>%  # ensure stfips is same type
  left_join(state_info, by = "stfips")
```

```{r}
# Add los_binned to your dataset
teds_d_completed <- teds_d_completed %>%
  mutate(
    los_binned = case_when(
      los == 1 ~ "1 day",
      los >= 2 & los <= 10 ~ "2-10 days",
      los >= 11 & los <= 21 ~ "11-21 days",
      los >= 22 & los <= 30 ~ "22-30 days",
      los == 31 ~ "31-45 days",
      los == 32 ~ "46-60 days",
      los == 33 ~ "61-90 days",
      los == 34 ~ "91-120 days",
      los == 35 ~ "121-180 days",
      los == 36 ~ "181-365 days",
      los == 37 ~ ">365 days",
      TRUE ~ NA_character_
    ),
    los_binned = factor(
      los_binned,
      levels = c("1 day", "2-10 days", "11-21 days", "22-30 days",
                 "31-45 days", "46-60 days", "61-90 days",
                 "91-120 days", "121-180 days", "181-365 days", ">365 days"),
      ordered = FALSE
    )
  )

```


```{r}
teds_d_completed <- teds_d_completed %>%
  mutate(
    hlthins = as.character(hlthins),  # Ensure all values are character
    hlthins = case_when(
      hlthins == "1" ~ "Private insurance, Blue Cross/Blue Shield, HMO",
      hlthins == "2" ~ "Medicaid",
      hlthins == "3" ~ "Medicare, other (e.g. TRICARE, CHAMPUS)",
      hlthins == "4" ~ "None",
      TRUE ~ hlthins  # Pass through any unmatched values as strings
    )
  )

```



```{r}
state_hlthins_hist_plot  <- ggplot(teds_d_completed, aes(x = los_binned, fill = as.factor(hlthins))) +
  geom_bar(color = "black", position = "stack") +
  facet_wrap(~State, scales = "free_y") +
  scale_fill_viridis_d(direction = -1) +
  theme_minimal() +
  labs(x = "Length of Stay (Binned)", fill = "Health Insurance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.8, size = 6))

state_hlthins_hist_plot

```


```{r}
ggsave(
  filename = here("plots", "state_hlthins_hist_plot.png"),
  plot = state_hlthins_hist_plot,
  device = "png",
  width = 16,  # Adjust width
  height = 9,  # Adjust height
  dpi = 300
)
```

